### 2015-07-23 chapters/ipa_meets_unicode.tex
readability edits

The International Phonetic Alphabet (IPA) is a common standard in linguistics to
transcribe sounds of spoken language into discrete segments using a Latin-based 
alphabet. Although IPA is reasonably easily adhered to with pen and paper, it is not trivial to encode
IPA characters electronically. Similar to the previous chapter, in this chapter
we discuss various pitfalls with the encoding of IPA in the Unicode Standard.\@ We will
specifically refer to the 2005 version of the IPA \citep{IPA2005} and the 7.0 
version of Unicode \citep{Unicode2014}.

@@

There are several pitfalls to be aware of when using the Unicode Standard to
encode IPA.\@ As we have said before, from a linguistic perspective it might
sometimes look like the Unicode Consortium is making incomprehensible decisions,
but it is important to realize that the consortium has tried and is continuing
to try to be as consistent as possible across a wide range of use cases, and it
does place linguistic traditions above other orthographic possibilities. Furthermore, 
when we look at the history of how the IPA meet Unicode, we see that many of the 
decisions for IPA symbols in the Unicode Standard come directly from the 
International Phonetic Association itself. Therefore as users, many pitfalls that we 
encounter have their grounding in the historical principles of the IPA, as 
well as the technological considerations involved in creating a single 
multilingual computing encoding.

In general, we strongly suggest to linguists to not complain about any decisions in
the Unicode Standard, but to try and understand the rationale of the International Phonetic Association 
and the Unicode Consortium (which in our experience is almost always well-conceived) and devise
ways to work with any unexpected behavior. Many of the current problems derive
from the fact that the IPA is clearly historically based on the Latin script,
but different enough from most other Latin-based writing systems to warrant
special attention. This ambivalent status of the IPA glyphs (partly Latin,
partly special) is unfortunately also attested in the treatment of IPA in the
Unicode Standard. In retrospect, it might have been better to consider the IPA
(and other transcription systems) to be a special kind of script within
the Unicode Standard, and treat the obvious similarity to Latin glyphs as a
historical relic. All IPA glyphs would then have their own code points, instead
of the current situation in which some IPA glyphs have special code points,
while others are treated as being identical to the regular Latin characters.
Yet, the current situation, however unfortunate, is unlikely to change, so as
linguists we must learn to deal with the specific pitfalls of IPA within
the Unicode Standard.

@@

Established in 1886, the International Phonetic Association (henceforth Association) 
has long maintained a standard alphabet, the International Phonetic Alphabet (IPA),
\footnote{Also referred to as API, for \textit{Association Phonétique Internationale}.} 
which is a common standard in linguistics to transcribe sounds of spoken languages. 
It was first published in 1888 as an international system of phonetic transcription 
for oral languages and for pedagogical purposes. The draft of 1887 contained phonetic 
values for English, French and German. Diacritics for length and nasalization were 
already present in this draft and the same symbols are still used today.

Although the original IPA alphabet was a list of symbols with pronunciation examples 
from words in different languages, in 1900 the symbols were organized into chart and 
given phonetic feature labels for consonants (e.g.\ for manner of articulation: `plosives', 
`nasales', `fricatives'; for place of articulation: `bronchials', `laryngeales', `labiales') 
and for vowels (`fermeées', `mi-fermeées', `mi-overtes', `overtes'). Over the last century 
or so, the structure of the chart has changed with increases in phonetic knowledge, and 
thus like notational systems in other scientific disciplines, the IPA reflects facts and 
theories (of phonetic knowledge) that have developed over time. It is natural then that 
the IPA is modified occasionally to accommodate scientific innovations and discoveries. 
(In fact, this is part of the Association's mandate.) These changes are captured in 
the revisions to the IPA.\footnote{For a detailed history see: 
\url{https://en.wikipedia.org/wiki/History\_of\_the\_International\_Phonetic_Alphabet}}

Over the years there has been several revisions, many minor. For example, 
articulation labels -- what we now call features even though the IPA avoids this term -- 
have changed (e.g.\ `lips', `throat', `rolled'). Phonetic symbol values have changed 
(e.g.\ voiceless is no longer marked by <h>). Symbols have been dropped (e.g.\ the 
caret diacritic denoting `long and narrow' is longer used). And many symbols have 
been added to reflect contrastive sounds found in the world's very diverse 
phonological systems.

@@

Third, the IPA has special markers for velarization <\dia{02E0}> and
pharyngealization <\dia{02E4}>. Confusingly, there is also a marker for
``velarized or pharyngealized'', using the <\dia{0334}> \textsc{combining tilde
overlay} at \uni{0334}. The tilde overlay seems to be extremely rarely used. We 
suggest to try and avoid using the tilde overlay, though for reasons of backward 
compatibility we will allow it in strict-IPA.\@

@@

% length IPA 1999:22
% "Note: as in Finnish orthography, length can also be indicated in phonetic transcription by double letters: e.g . Finnish maatto [rncctto] 'electrical earth/ground'."

% syllable breaks and word boundaries <\s, ., |, ||, tie-bar below>
% White spaces can be used to indicate word boundaries. Syllable breaks can be marked when required. The other two boundary symbols are used to mark the domain of larger prosodic units. There is also a linking symbol that can be used for explicitly indicating the lack of a boundary.

%  (In all these transcriptions, no spaces between words have been included. This is inevitable in an impressionistic transcription where it is not yet known how the utterance divides into words. In phonemic and allophonic transcriptions it is common to include spaces to aid legibility, but their theoretical validity is problematic.)

@@

% Wells1995:10 "The underscore could in principle also be pressed into service to represent the IPA tie bar. The current chart mentions its use only for affricates and double articulations, and then only "if necessary"."

In the major revision of the IPA in 1932, affricates were represented by two consonants 
(e.g.\ <tʃ>), or ligatures <ʧ>, or with the tie-bar <t͡ʃ>. In the 1938 revision the tie-bar's 
semantics were broadened to indicate simultaneous articulation, e.g.\ in labial velars 
such as <k͡p>. Thus the tie-bar is a convenient diacritic for visually tokenizing input 
strings into chunks of phonetically salient groups, including affricates, doubly articulated 
consonants or diphthongs.

The tie bar can be placed above or below the base characters, e.g. <\charis{t͡s}>
or <\charis{t͜s}>.\footnote{Note that the \textsc{undertie} at \uni{203F} symbol 
looks like and is easily confused with the tie-bar below. However, it is a different 
character and has a different function in IPA. The undertie is used as a linking 
symbol to indicate the lack of a boundary, e.g.\ petit ami [pətit‿ami] `boyfriend' 
($<$ French).} IPA allows both options. The choice between the two symbols
is purely for legible rendering; there is no difference in semantics between the
two symbols. However, rendering is such a problematic issue for tie bars in
general that many linguists simply do not use them. Just looking at a few
different fonts already indicates that actually no font designer really gets the
placement right in combination with superscripts and subscripts. If really
necessary, we propose to standardize on the tie bar above the base characters,
using a \textsc{combining double inverted breve} at \uni{0361}.

@@

% TODO: link back to discussion on IPA principles and how ligatures got brought along historically

@@

There turn out to be a lot of characters in the IPA that could be conceived as
using any of these elements, like <ɲ>, <ɳ>, <ɨ> or <ø>. However, all such
characters exist as well as precomposed combination in Unicode, and these
precomposed characters should preferably be used.\footnote{The IPA does not
describe any character for a voiced retroflex implosive, which would
transparently be \charis{ᶑ}. We propose to add this character to the IPA, see
Section~\ref{ipa-additions}.} When instead combinations of a base character with
diacritic are used, then these combinations are not canonical equivalent to the
precomposed combinations. This means that any search will not find both at the
same time.

A similar problem arises with the rhotic hook. There are two precomposed
characters in Unicode with a rhotic hook, which are not canonical equivalent 
with a combination of the vowel with a separately encoded hook:
\begin{itemize}
  \item[] <ɚ> \textsc{latin small letter schwa with hook} at \uni{025A}
  \item[] <ɝ> \textsc{latin small letter reversed open e with hook} at \uni{025D}
\end{itemize}
All other combinations of vowels with rhotic hooks will have to be made by using
<\dia{02DE}> \textsc{modifier letter rhotic hook} at \uni{02DE}, because there
is not complete set of precomposed characters with rhotic hooks. For that reason
we propose to not use the two precomposed characters with hooks mentioned above,
but always use the separate rhotic hook at \uni{02DE} in IPA.\@

A similar situation arises with <\dia{0334}> \textsc{combining tilde overlay} at
\uni{0334}. The main reason some phoneticians like to use this in IPA is to mark
the ``dark'' <l> in English codas, using the character <ɫ> \textsc{latin small
letter l with middle tilde} at \uni{026B}. This character is not canonically
equivalent to the combination <l> + <\dia{0334}>, so one of the two possible
encodings has to be chosen. Because the tilde overlay is described as a general
mechanism by the IPA, we propose to use the separated <\dia{0334}>
\textsc{combining tilde overlay} at \uni{0334}. However, note that phonetically 
this seems to be (almost) superfluous (see Section~\ref{pitfall-multiple-options-ipa}) 
and the typical usage in the form of <ɫ> is (almost) a homoglyph with <ɬ> (see 
Section~\ref{pitfall-homoglyphs-in-IPA}). For these reasons we suggest to try 
and avoid the tilde overlay.

@@

Although linguists might expect these characters to belong together with the
character in front of them, at least for <ʰ> \textsc{modifier letter small h} at
\uni{02B0} the Unicode Consortium's decision to treat it as a separate character
is also linguistically correct, because according to the IPA it can be used both
for aspiration (more precisely post-aspiration following the base character) and
pre-aspiration (preceding the base character). The default combination of Spacing Modifiers with
the preceding character can be specified in orthography profiles (see
Chapter~\ref{orthography-profiles}).

@@

There is one aspect of ordering for which the Unicode Standard does present a
canonical solution. However, it is uncontroversial from a linguistic
perspective. Diacritics in the Unicode Standard (i.e.~Combining Diacritical
Marks, see above) are classified in Canonical Combining Classes. In practice,
the diacritics are distinguished by their position relative to the base
character.\footnote{See
\url{http://unicode.org/reports/tr44/\#Canonical\_Combining\_Class\_Values} for a
detailed description.} When applying a Unicode normalization (NFC or NFD, see
Section~\ref{pitfall-canonical-equivalence}), the diacritics in different
positions are put in a specified order. This process therefore harmonizes the
difference between different encodings, e.g.\ of an extra-short creaky voice
vowel <ḛ̆>. This grapheme cluster can be encoded either as
<e>+<\dia{0306}>+<\dia{0330}> or as <e>+<\dia{0330}>+<\dia{0306}>. To
prevent this twofold encoding, the Unicode Standard specifies the second
ordering as canonical (in this case, diacritics below are put before diacritics
above).

When encoding a string according to the Unicode Standard, it is possible to do
this either using the NFC (composition) or NFD (decomposition) normalization.
Decomposition implies that precomposed characters (like <á> \textsc{latin small
letter a with acute} at \uni{00E1}) will be split into its parts. This might
sound preferable for a linguistic analysis, as the different diacritics are
separated from the base characters. However, note that most attached elements
like strokes (e.g.~in the <ɨ>), retroflex hooks (e.g.~in <ʐ>) or rhotic hooks
(e.g.~in <ɝ>) will not be decomposed, but strangely enough a cedilla (like in
<ç>) will be decomposed (see Section~\ref{pitfall-missing-decomposition}). In
general, Unicode decomposition does not behave like a feature decomposition as
expected from a linguistic perspective. It is thus important to consider Unicode
decomposition only as a technical procedure, and not assume that it is
linguistically sensible.

@@

By Unicode Canonical Combining Classes, the diacritics on top of a character,
like <\dia{0334}> (Combining Class number 1), always come before diacritics
below (Combining Class number 220), which in turn always come before diacritics
above (Combining Class number 230), which in turn come before diacritics over
multiple characters like the tie bar <\dia{0361}{\large\fontspec{CharisSIL}◌}>
(Combining Class number 233). We follow this order, but add the other IPA
diacritics (which are not diacritics in the Unicode sense) between diacritics
below and the tie bar. Within all these classes of diacritics there is no canonical
ordering specified by Unicode, so we propose an explicit ordering here.

@@

With each revision of the IPA, many decisions need to be made by 
the Association as to which symbols should be added, removed or 
changed. (Perhaps that's why the IPA is updated infrequently.) 
For example, in the 1989 revision of the IPA at the Kiel Convention, 
changes to specific symbols (in previous charts) were debated and 
the Association's members certain decisions. The prevailing mood at 
the convention was not to change specific symbols unless a strong 
case was made \cite{Ladefoged1990a}. For example, two such decisions 
included \cite[62]{Ladefoged1990a}:

\begin{itemize}
	\item symbols for clicks were changed from < ʇ, ʖ, ʗ > and replaced by < ǀ, ǁ, ǃ > because the latter were the symbols used by nearly all Khoisanists and Bantuists
	\item the Americanist tradition of < š, ž, č, ǰ > were not adopted because the Association members at the convention ``were not sufficiently impressed by arguments ... to the effect that these sounds formed a natural class, and thus is would be appropriate to recognize this by maintaining a common aspect to their symbolism''
\end{itemize}

>>>>

\section{Introduction}

The International Phonetic Alphabet (IPA) is a common standard in linguistics to
transcribe sounds of spoken language into discrete segments using a Latin-based
alphabet. Although IPA is reasonably easily adhered to with pen and paper, it is
not trivial to encode IPA characters electronically. Similar to the previous
chapter, in this chapter we discuss various pitfalls with the encoding of IPA in
the Unicode Standard. We will specifically refer to the 2005 version of the
IPA \citep{IPA2005} and the 7.0 version of Unicode \citep{Unicode2014}.

@@

There are several pitfalls to be aware of when using the Unicode Standard to
encode IPA.\@ As we have said before, from a linguistic perspective it might
sometimes look like the Unicode Consortium is making incomprehensible decisions,
but it is important to realize that the consortium has tried and is continuing
to try to be as consistent as possible across a wide range of use cases, and it
does place linguistic traditions above other orthographic choices. Furthermore,
when we look at the history of how the IPA met Unicode, we see that many of the
decisions for IPA symbols in the Unicode Standard come directly from the
International Phonetic Association itself. Therefore, many pitfalls that we will
encounter have their grounding in the history of the principles of the IPA, as
well as in the technological considerations involved in creating a single
multilingual encoding. In general, we strongly suggest to linguists to not
complain about any decisions in the Unicode Standard, but to try and understand
the rationale of the International Phonetic Association and the Unicode
Consortium (both of which are almost always well-conceived in our experience)
and devise ways to work with any unexpected behavior.

Many of the current problems derive from the fact that the
IPA is clearly historically based on the Latin script, but different enough from
most other Latin-based writing systems to warrant special attention. This
ambivalent status of the IPA glyphs (partly Latin, partly special) is
unfortunately also attested in the treatment of IPA in the Unicode Standard. In
retrospect, it might have been better to consider the IPA (and other
transcription systems) to be a special kind of script within the Unicode
Standard, and treat the obvious similarity to Latin glyphs as a historical
relic. All IPA glyphs would then have their own code points, instead of the
current situation in which some IPA glyphs have special code points, while
others are treated as being identical to the regular Latin characters. Yet, the
current situation, however unfortunate, is unlikely to change, so as linguists
we must learn to deal with the specific pitfalls of IPA within the Unicode
Standard.

@@

Established in 1886, the International Phonetic Association (henceforth Association) 
has long maintained a standard alphabet, the International Phonetic Alphabet (IPA),
%\footnote{Also referred to as API, for \textit{Association Phonétique Internationale}.} 
which is a common standard in linguistics to transcribe sounds of spoken
languages. It was first published in 1888 as an international system of phonetic
transcription for oral languages and for pedagogical purposes. The draft of 1887
contained phonetic values for English, French and German. Diacritics for length
and nasalization were already present in this draft and the same symbols are
still used today.

Originally, the IPA alphabet was a list of symbols with pronunciation examples
from words in different languages. In 1900 the symbols were first organized into
chart and were given phonetic feature labels for consonants (e.g.\ for manner of
articulation: `plosives', `nasales', `fricatives'; for place of articulation:
`bronchials', `laryngeales', `labiales') and for vowels (e.g.\ `fermeées',
`mi-fermeées', `mi-overtes', `overtes'). Throughout the last centory, the
structure of the chart has changed with increases in phonetic knowledge, and
thus, like notational systems in other scientific disciplines, the IPA reflects
facts and theories (of phonetic knowledge) that have developed over time. It is
natural then that the IPA is modified occasionally to accommodate scientific
innovations and discoveries. (In fact, this is part of the Association's
mandate.) These changes are captured in the revisions to the IPA.
%\footnote{For a detailed history see:
%\url{https://en.wikipedia.org/wiki/History\_of\_the\_International\_Phonetic_Alphabet}}

Over the years there has been several revisions, many minor. For example,
articulation labels -- what often are call features even though the IPA avoids
this term -- have changed (e.g.\ terms like `lips', `throat' or `rolled' are not
used anymore). Phonetic symbol values have changed (e.g.\ voiceless is no longer
marked by <h>). Symbols have been dropped (e.g.\ the caret diacritic denoting
`long and narrow' is no longer used). And many symbols have been added to reflect
contrastive sounds found in the world's very diverse phonological systems.

@@

Third, the IPA has special markers for velarization <\dia{02E0}> and
pharyngealization <\dia{02E4}>. Confusingly, there is also a marker for
``velarized or pharyngealized'', using the <\dia{0334}> \textsc{combining tilde
overlay} at \uni{0334}. The tilde overlay seems to be extremely rarely used. We 
suggest to try and avoid using the tilde overlay, though for reasons of backward 
compatibility we will allow it in valid-IPA.\@

@@

% length IPA 1999:22
% "Note: as in Finnish orthography, length can also be indicated in phonetic
% transcription by double letters: e.g . Finnish maatto [rncctto] 'electrical
% earth/ground'."

% syllable breaks and word boundaries <\s, ., |, ||, tie-bar below>
% White spaces can be used to indicate word boundaries. Syllable breaks can be
% marked when required. The other two boundary symbols are used to mark the domain
% of larger prosodic units. There is also a linking symbol that can be used for
% explicitly indicating the lack of a boundary.

% (In all these transcriptions, no spaces between words have been included. This
% is inevitable in an impressionistic transcription where it is not yet known how
% the utterance divides into words. In phonemic and allophonic transcriptions it
% is common to include spaces to aid legibility, but their theoretical validity is
% problematic.)

@@

% Wells1995:10 "The underscore could in principle also be pressed into service
% to represent the IPA tie bar. The current chart mentions its use only for
% affricates and double articulations, and then only "if necessary"."

In the major revision of the IPA in 1932, affricates were represented by two
consonants <tʃ>, ligatures <ʧ>, or with the tie-bar <t͡ʃ>. In the
1938 revision the tie-bar's semantics were broadened to indicate simultaneous
articulation, as for example in labial velars such as <k͡p>. Thus, the tie-bar is a
convenient diacritic for visually tokenizing input strings into chunks of
phonetically salient groups, including affricates, doubly articulated consonants
or diphthongs.

The tie bar can be placed above or below the base characters, e.g.
<\charis{t͡s}> or <\charis{t͜s}>. IPA allows both options. The choice between
the two symbols is purely for legible rendering; there is no difference in
semantics between the two symbols. However, rendering is such a problematic
issue for tie bars in general that many linguists simply do not use them. Just
looking at a few different fonts already indicates that actually no font
designer really gets the placement right in combination with superscripts and
subscripts. If really necessary, we propose to standardize on the tie bar above
the base characters, using a \textsc{combining double inverted breve} at
\uni{0361}.\footnote{Also note that the \textsc{undertie} at \unif{203F} looks
like the tie bar below and is easily confused with it. However, it is a
different character and has a different function in IPA. The undertie is used as
a linking symbol to indicate the lack of a boundary, e.g.\ French \textit{petit
ami} [pətit\charis{‿}ami] `boyfriend'.}

@@

% TODO: link back to discussion on IPA principles and how ligatures got brought
% along historically

@@

There turn out to be a lot of characters in the IPA that could be conceived as
using any of these elements, like <ɲ>, <ɳ>, <ɨ> or <ø>. However, all such
characters exist as well as precomposed combination in Unicode, and these
precomposed characters should preferably be used. When instead combinations of a base character with
diacritic are used, then these combinations are not canonical equivalent to the
precomposed combinations. This means that any search will not find both at the
same time.

% \footnote{The IPA does not
% describe any character for a voiced retroflex implosive, which would
% transparently be \charis{ᶑ}. We propose to add this character to the IPA, see
% Section~\ref{ipa-additions}.}

A similar problem arises with the rhotic hook. There are two precomposed
characters in Unicode with a rhotic hook, which are not canonical equivalent 
with a combination of the vowel with a separately encoded hook:
\begin{itemize}
  \item[] <ɚ> \textsc{latin small letter schwa with hook} at \uni{025A}
  \item[] <ɝ> \textsc{latin small letter reversed open e with hook} at \uni{025D}
\end{itemize}
All other combinations of vowels with rhotic hooks will have to be made by using
<\dia{02DE}> \textsc{modifier letter rhotic hook} at \uni{02DE}, because there
is no complete set of precomposed characters with rhotic hooks in Unicode. For
that reason we propose to not use the two precomposed characters with hooks
mentioned above, but always use the separate rhotic hook at \uni{02DE} in IPA.\@

A similar situation arises with <\dia{0334}> \textsc{combining tilde overlay} at
\uni{0334}. The main reason some phoneticians like to use this in IPA is to mark
the ``dark'' <l> in English codas, using the character <ɫ> \textsc{latin small
letter l with middle tilde} at \uni{026B}. This character is not canonically
equivalent to the combination <l>~+~<\dia{0334}>, so one of the two possible
encodings has to be chosen. Because the tilde overlay is described as a general
mechanism by the IPA, we propose to use the separated <\dia{0334}>
\textsc{combining tilde overlay} at \uni{0334}. However, note that phonetically 
this seems to be (almost) superfluous (see Section~\ref{pitfall-multiple-options-ipa}) 
and the typical usage in the form of <ɫ> is (almost) a homoglyph with <ɬ> (see 
Section~\ref{pitfall-homoglyphs-in-IPA}). For these reasons we also suggest to try 
and avoid the tilde overlay completely.

@@

Although linguists might expect these characters to belong together with the
character in front of them, at least for <ʰ> \textsc{modifier letter small h} at
\uni{02B0} the Unicode Consortium's decision to treat it as a separate character
is also linguistically correct, because according to the IPA it can be used both
for post-aspiration (following the base character) and pre-aspiration (preceding
the base character). The default combination of Spacing Modifiers with the
preceding character can be specified in orthography profiles (see
Chapter~\ref{orthography-profiles}).

@@

There is one (minor) aspect of ordering for which the Unicode Standard does
present a canonical solution. Fortunately, this is uncontroversial from a
linguistic perspective. Diacritics in the Unicode Standard (i.e.~Combining
Diacritical Marks, see Section~\ref{pitfall-different-notions-of-diacritics})
are classified in so-called \textsc{canonical combining classes}. In practice,
the diacritics are distinguished by their position relative to the base
character.\footnote{See
\url{http://unicode.org/reports/tr44/\#Canonical\_Combining\_Class\_Values} for
a detailed description.} When applying a Unicode normalization (NFC or NFD, see
Section~\ref{pitfall-canonical-equivalence}), the diacritics in different
positions are put in a specified order. This process therefore harmonizes the
difference between different encodings in some situations, for example in the
case of an extra-short creaky voice vowel <ḛ̆>. This grapheme cluster can be
encoded either as <e>+<\dia{0306}>+<\dia{0330}> or as
<e>+<\dia{0330}>+<\dia{0306}>. To prevent this twofold encoding, the Unicode
Standard specifies the second ordering as canonical (namely, diacritics
below are put before diacritics above).

% The next paragraph does not belong here - we should put it somewhere else

When encoding a string according to the Unicode Standard, it is possible to do
this either using the NFC (composition) or NFD (decomposition) normalization (see
Section~\ref{pitfall-canonical-equivalence}).
Decomposition implies that precomposed characters (like <á> \textsc{latin small
letter a with acute} at \uni{00E1}) will be split into its parts. This might
sound preferable for a linguistic analysis, as the different diacritics are
separated from the base characters. However, note that most attached elements
like strokes (e.g.~in the <ɨ>), retroflex hooks (e.g.~in <ʐ>) or rhotic hooks
(e.g.~in <ɝ>) will not be decomposed, but strangely enough a cedilla (like in
<ç>) will be decomposed (see Section~\ref{pitfall-missing-decomposition}). In
general, Unicode decomposition does not behave like a feature decomposition as
expected from a linguistic perspective. It is thus important to consider Unicode
decomposition only as a technical procedure, and not assume that it is
linguistically sensible.

@@

By Unicode Canonical Combining Classes, the diacritics on top of a character,
like <\dia{0334}> (Combining Class number 1), always come before diacritics
below (Combining Class number 220), which in turn always come before diacritics
above (Combining Class number 230), which in turn come before diacritics over
multiple characters like the tie bar <\dia{0361}{\large\fontspec{CharisSIL}◌}>
(Combining Class number 233). We follow this order, but add the other IPA
diacritics (which are not diacritics in the Unicode sense) between diacritics
below and the tie bar. Further, within all these classes of diacritics there is
no canonical ordering specified by Unicode, so we propose an explicit ordering
here.

@@

With each revision of the IPA, many decisions need to be made by 
the Association as to which symbols should be added, removed or 
changed. For example, in the 1989 revision of the IPA at the Kiel Convention, 
changes to specific symbols (in previous charts) were debated and 
the Association's members certain decisions. The prevailing mood at 
the convention was not to change specific symbols unless a strong 
case was made \cite{Ladefoged1990a}. For example, two such decisions 
included:

\begin{itemize}
	\item symbols for clicks were changed from <ʇ~ʖ~ʗ> to <ǀ~ǁ~ǃ>
       because the latter were the symbols used by nearly all Khoisanists and
       Bantuists.
	\item the Americanist tradition of using using <\dia{030C}>, a
       \textsc{combining caron} at \uni{030C} for all post-alveolar sounds, like
       in <š~ž~č~ǰ>, was not adopted because the Association members at the
       convention ``were not sufficiently impressed by arguments ... to the
       effect that these sounds formed a natural class, and thus is would be
       appropriate to recognize this by maintaining a common aspect to their
       symbolism'' \citep[62]{Ladefoged1990a}. 
\end{itemize}
